---
title: "Class 2 — Reproducible Single-Cell and Spatial Analysis with R and Python"
subtitle: "From raw data to biological insight"
---

## Introduction to Spatial Transcriptomics

---

## 1. Questions only ST answers

| Question | Example |
|---|---|
| Where is a cell type located? | Immune cells at the tumor border |
| Who neighbors whom? | Niches, microenvironment |
| How does expression change along an axis? | Cortex layers, intestinal crypt |
| Which ligand receptor pairs are spatially plausible? | Cells in physical contact |
| Which tissue domains exist by expression, not morphology? | Unsupervised domain discovery |

---

## 2. The main families

| | Sequencing based (seq-based) | Imaging based (in situ) | Region based (ROI) |
|---|---|---|---|
| **Principle** | Slide carries oligos with a known positional barcode. mRNA diffuses, is captured, sequenced, and position is recovered from the barcode | Fluorescent probes hybridize to transcripts inside intact tissue. Each gene gets a color code over successive rounds | User selects regions on the image; only those are collected and sequenced |
| **Genes** | Untargeted, whole transcriptome | Targeted, closed panel | Whole transcriptome or panels |
| **Strength** | Discovery without prior hypothesis | Subcellular resolution, high sensitivity | Good for defined anatomical regions |
| **Intrinsic limit** | Poly-A capture and lateral diffusion limit resolution and sensitivity | You must know the genes beforehand | No continuous map |
| **Examples** | Visium, Visium HD, Slide-seq, Stereo-seq | Xenium, CosMx, MERSCOPE, seqFISH, STARmap | GeoMx DSP, LCM-seq, Tomo-seq |

Imaging based has two subfamilies: **multiplexed ISH** (MERFISH, seqFISH, Xenium, CosMx, MERSCOPE) and **ISS, in situ sequencing** (STARmap, BaristaSeq, HybISS), where the transcript is sequenced on the slide.

---

## 3. Platform comparison

Approximate values. Confirm current vendor specifications before designing an experiment.

### 3.1 Sequencing based

| Platform | Resolution | Single cell? | Genes | Sample | Notes |
|---|---|---|---|---|---|
| Original ST (Ståhl 2016) | 100 µm | No | Whole | FF | Basis of Visium |
| **10x Visium** | 55 µm spots, 100 µm pitch, ~5k spots | No, ~1 to 10 cells per spot | Whole | FF and FFPE (CytAssist) | Most used. Area 6.5 x 6.5 mm |
| **10x Visium HD** | 2 µm bins (output at 8 µm) | Near. A bin is not a cell | Whole | FFPE | Continuous surface, no gaps |
| Slide-seq / V2 | 10 µm beads | Near | Whole | FF | Academic. Lower sensitivity than Visium |
| **Stereo-seq (BGI)** | ~220 nm DNBs | Subcellular, needs binning or segmentation | Whole | FF | Centimeter scale field of view |
| HDST | 2 µm beads | Subcellular | Whole | FF | Academic, low efficiency |
| DBiT-seq | 10 to 50 µm pixels | No | Whole | FF | RNA + protein on the same tissue |
| Seq-Scope | Submicron | Subcellular | Whole | FF | Uses an Illumina flow cell |

### 3.2 Imaging based

| Platform | Resolution | Single cell? | Genes | Sample | Notes |
|---|---|---|---|---|---|
| smFISH | Subcellular | Yes | 1 to 4 | FF and FFPE | Gold standard for sensitivity, no multiplexing |
| **MERFISH / MERSCOPE (Vizgen)** | ~100 nm | After segmentation | ~140 to 1,000 | FF and FFPE | High sensitivity |
| seqFISH+ | Subcellular | After segmentation | Up to ~10,000 | FF | Academic, highest plex by ISH |
| **10x Xenium** | ~200 nm | After segmentation | ~250 to 500, up to ~5,000 (Prime) | FF and FFPE | Multiple cell boundary stains |
| **NanoString CosMx** | Subcellular | After segmentation | ~1,000 to 6,000, whole transcriptome versions | FF and FFPE | RNA + protein |
| ISS / HybISS / BaristaSeq | Subcellular | After segmentation | Tens to hundreds | FF and FFPE | Sequencing on the slide |
| STARmap | Subcellular, 3D | Yes | Hundreds | FF | Cleared tissue, 3D volume |

### 3.3 Region based

| Platform | Resolution | Single cell? | Genes | Sample | Notes |
|---|---|---|---|---|---|
| **GeoMx DSP** | ROI of ~10 to 600 µm | No, can enrich by marker | Whole (~18k) or panels | FF and FFPE | Not a continuous map |
| LCM-seq | Microdissected region | Depends on the cut | Whole | FF and FFPE | Labor intensive, low throughput |

---

## 4. Advantages and limitations

| | Seq-based | Imaging based |
|---|---|---|
| **Advantages** | Whole transcriptome, discovery. Detects unexpected genes. Workflow close to scRNA-seq. Low cost per gene. Visium has the largest user base and most validated tools | True subcellular resolution. High sensitivity. Real per cell data (niches and contacts without deconvolution). Keeps morphology, integrates with H&E and IF. Works well with FFPE and degraded RNA |
| **Limitations** | Spots mix cells, rare types get diluted. Needs **deconvolution** with a scRNA-seq reference. Low capture efficiency, sparse matrices. Poly-A only. Lateral diffusion blurs position. Permeabilization must be tuned per tissue. Small fixed area. High cost per sample. In Visium HD and Stereo-seq the **bin is not a cell** | **Closed panel**, no discovery outside it. **Segmentation is the main bottleneck**: errors create chimeric profiles and false cell types. Optical crowding of abundant genes. Autofluorescence (lipofuscin, collagen, red blood cells). Huge data (hundreds of GB to TB). Long instrument time, high cost. Less mature analysis ecosystem |

---

## 5. How to choose a platform

Answer in this order.

| # | Question | Answer | Direction |
|---|---|---|---|
| 1 | Do you know which genes to measure? | No, I want to discover | Seq-based: Visium, Visium HD, Stereo-seq |
| | | Yes, defined markers | Imaging: Xenium, CosMx, MERSCOPE |
| 2 | Do you need single cell resolution? | Yes, contact or neighborhood | Imaging |
| | | No, domains are enough | Visium |
| 3 | Which sample do you have? | Archival FFPE | Visium FFPE/CytAssist, Visium HD, Xenium, CosMx |
| | | Fresh frozen | Essentially all platforms |
| 4 | How much tissue area? | A few mm | Visium |
| | | Whole organ or embryo | Stereo-seq or imaging with many FOVs |
| 5 | Is there a scRNA-seq reference? | No | Avoid deconvolution: imaging or finer resolution |
| | | Yes, good quality | Visium + deconvolution is solid and cheap |

> **Common strategy in practice:** Visium or Visium HD for untargeted discovery, then validation with a Xenium or CosMx panel on the same cohort.

---

## 6. Glossary

| Term | Definition |
|---|---|
| **Spot / bead / bin** | Spatial capture unit in seq-based methods |
| **Spatial barcode** | Sequence that identifies where a transcript came from |
| **UMI** | Unique Molecular Identifier. Tags original molecules, corrects PCR duplication |
| **Deconvolution** | Estimating cell type proportions in a multicellular spot |
| **Segmentation** | Computationally delimiting each cell in the image |
| **Plex** | Number of genes measured at once |
| **SVG** | Spatially variable gene: expression varies non randomly in space |
| **Spatial domain** | Tissue region with a coherent transcriptional profile |
| **Niche** | Cell composition in the neighborhood of a cell |
| **FOV** | Field of view |
| **FF / FFPE** | Fresh frozen / Formalin fixed paraffin embedded |
| **Permeabilization** | Step that releases mRNA for capture. Critical in seq-based |

---

## 7. Single-cell vs. spatial: preprocessing and cell annotation

Both fields share the same conceptual pipeline — QC, normalization, dimensionality reduction, clustering, annotation. The difference is not the list of steps, it is **what a row of the count matrix actually represents**.

In scRNA-seq the cell is defined by the chemistry itself: the tissue is dissociated, one cell goes into one droplet, and every transcript in that droplet is by construction assigned to that cell. In spatial data the cell is never given. It has to be *reconstructed computationally*, and the way it is reconstructed depends entirely on the technology family.

### 7.1 What the measurement unit is

| Technology family | Unit in the matrix | Relation to a cell |
|---|---|---|
| scRNA-seq / snRNA-seq | Droplet or well | One cell (barring doublets and empty droplets) |
| **Spot-based** (Visium, original ST, Slide-seq) | **Spot or bead** | Fixed capture area. A 55 µm Visium spot typically pools ~1 to 10 cells of mixed types. The cell does not exist in the data |
| **High-resolution seq-based** (Visium HD, Stereo-seq, Seq-Scope) | **Bin** (arbitrary square, e.g. 2, 8 or 16 µm) | A bin is a grid square, not a biological object. It may cut one cell across several bins, or pool fragments of neighbors |
| **Imaging-based** (Xenium, CosMx, MERSCOPE) | **Segmented cell** | Raw data is a *point cloud of individual transcripts with (x, y) coordinates*. A "cell" only exists after segmentation draws a boundary around a group of those points |
| Region-based (GeoMx, LCM) | User-drawn ROI | Whatever the user selected. No single-cell claim at all |

> The consequence: in spot- and bin-based data the analytical problem is **mixture** (which cell types compose this unit, and in what proportion). In imaging-based data the problem is **assignment** (which cell does this transcript belong to). These are different problems and they need different tools.

### 7.2 Preprocessing, step by step

| Step | scRNA-seq | Spatial |
|---|---|---|
| **QC metrics** | Counts per cell, genes per cell, % mitochondrial, doublet score | Same metrics, but the thresholds mean something different. A low-count Visium spot may be a genuinely sparse tissue region, not a dying cell. In imaging-based data, counts per cell scale with the panel size (a 300-gene panel gives far fewer counts than a whole transcriptome) and with cell area, so a "low-count" cell may simply be a small lymphocyte |
| **% mitochondrial** | Standard proxy for dying cells | Often unusable: mitochondrial genes are frequently **absent from targeted panels**, and in FFPE the metric is not interpretable the same way. Needs different QC proxies |
| **Doublet detection** | Core step (Scrublet, DoubletFinder) | Meaningless in spot/bin data — every spot is a "multiplet" by design; the problem is handled by deconvolution instead. In imaging-based data, "doublets" are not chemistry artifacts but **segmentation errors** (two cells merged into one), which is a different failure mode and needs different diagnostics |
| **Ambient RNA** | Free-floating RNA in the droplet suspension (SoupX, CellBender) | The analogous problem is **transcript spillover / lateral diffusion**: RNA that physically drifted away from its cell of origin, or extracellular transcripts sitting between cells. Imaging platforms include **negative control probes** (blank codewords) to estimate the false-detection rate directly |
| **Segmentation** | Not applicable — cells are already physically separated | The central preprocessing step for imaging-based data. See 7.3 |
| **Deconvolution** | Not applicable | Required for spot-based data to estimate cell-type proportions per spot (RCTD, Cell2location, SPOTlight, Stereoscope, cell2location) |
| **Normalization** | Library size + log, or SCTransform | Same formulas, but total counts per unit confound with **cell/spot area and local tissue density**, not only with capture efficiency. A dense lymphoid region will have higher counts per spot than sparse stroma for purely anatomical reasons — this is biology-driven, not technical, and normalizing it away can erase real signal |
| **Batch / slide effects** | Between samples and chemistries | Also between **slides, capture areas and fields of view (FOVs)** within the same experiment. FOV-edge artifacts are a known imaging-platform issue |
| **Spatial-specific QC** | None | Detecting which spots are actually under tissue, alignment/registration to the H&E or IF image, checking panel coverage for the cell types of interest, and inspecting the image itself for folds, tears and detachment |

### 7.3 Why segmentation is *the* problem in imaging-based data

**The raw data contains no cells.** A Xenium or CosMx run outputs a list of decoded transcripts, each with a gene identity and an (x, y) — often (x, y, z) — coordinate. Nothing in that file says which transcript came from which cell. Building the count matrix means drawing boundaries on the image and summing the transcripts that fall inside each boundary. **Everything downstream is therefore conditional on the quality of the cell-boundary detection.** If the boundary is wrong, the expression profile is wrong, and no amount of clustering, normalization or annotation will recover it.

**How the errors manifest**

| Error | Mechanism | Downstream consequence |
|---|---|---|
| **False merge** | Two adjacent cells enclosed in one boundary | A **chimeric profile**: a T cell fused with an epithelial cell produces a unit expressing both *CD3E* and *EPCAM*. It clusters separately and is easily mistaken for a novel "intermediate" or "transitional" cell state |
| **False split** | One cell divided into two units | Halves the counts, dilutes markers, creates artificially low-quality cells that get discarded by QC — biasing against large or irregularly shaped cell types |
| **Transcript spillover** | Transcripts of cell A assigned to cell B | The most insidious error: it does not look broken. Highly expressed genes from an abundant neighbor bleed into every surrounding cell, generating **false co-expression** and inflating apparent cell-cell "interaction" signal |
| **Orphan transcripts** | Real transcripts with no assignable cell (e.g. the nucleus lies outside the section plane) | Lost signal, systematically biased toward cell types with large or elongated morphology |

The critical point is that **these errors are indistinguishable from biology after the fact**. A chimeric merged cell looks exactly like a real hybrid cell state. Spillover looks exactly like a real ligand-receptor neighborhood. This is why segmentation must be inspected on the image, not merely accepted as a preprocessing default.

**Why nucleus expansion and Voronoi tessellation are weak solutions**

Both are geometric shortcuts used because the nuclear stain (DAPI) is trivially easy to segment while the *cell boundary* is not. Both replace a biological boundary with a mathematical one.

*Fixed-radius nucleus expansion* — segment nuclei, then dilate each by a constant distance (commonly 5–15 µm) and call that the cell:

- It assumes **every cell has the same size and shape**. A resting lymphocyte (~7 µm) and a macrophage, tumor cell or neuron (30 µm and up, with processes) get the same treatment. Small cells are over-expanded into their neighbors; large cells are truncated, losing most of their cytoplasmic transcripts.
- Most mRNA is **cytoplasmic**, and cytoplasmic geometry is exactly what the nuclear stain does not tell you. Polarized epithelium, elongated fibroblasts, dendritic processes and neuronal projections are all badly approximated by a disc.
- It fails hardest precisely where the biology is most interesting: **densely packed tissue** (lymph node, germinal centre, tumor nest), where every expansion necessarily crosses into a neighbor.

*Voronoi / Delaunay tessellation* — partition the plane so each point belongs to the nearest nucleus:

- It **partitions all available space**, so every transcript is assigned to some cell. There is no "unassigned", no extracellular compartment, no acellular matrix. Ambient and extracellular transcripts are forced into cells by construction.
- It produces **convex polygons that tile the plane**, which no real tissue does. Non-convex, interdigitating and wrapping morphologies (podocytes, astrocytes, dendritic cells) cannot be represented.
- Cells whose nucleus is **outside the plane of section** have no seed at all, so their territory is silently absorbed by neighbors — a 2D slice through a 3D tissue guarantees this happens.

The shared failure of both: they use only nuclear position, ignoring the transcript distribution itself and any real membrane information.

**What actually improves segmentation**

| Strategy | What it adds |
|---|---|
| **Membrane and boundary protein staining** | The direct fix: image the actual cell border. Xenium's multi-tissue stain (e.g. ATP1A1, E-cadherin, CD45) and CosMx's morphology markers (PanCK, CD45, CD298/B2M, membrane markers) provide channels that mark membranes and compartments, so the boundary is measured rather than assumed. Choosing markers that cover *your* cell types matters — a panel with only epithelial and immune boundary markers will still segment stroma poorly |
| **Deep-learning segmentation on those channels** | Cellpose, StarDist, Mesmer/DeepCell learn irregular, non-convex shapes from image data instead of imposing a disc or polygon. Only as good as the input stains, and usually benefits from tissue-specific fine-tuning |
| **Transcript-aware / probabilistic segmentation** | Baysor, ClusterMap, JSTA, BIDCell, proseg use the **transcript point cloud itself** as evidence: transcripts with coherent expression identity are grouped together, boundaries follow molecular composition rather than geometry alone, and — crucially — transcripts can be left **unassigned** instead of being forced into a cell. Combining these with a membrane-stain prior generally outperforms either alone |
| **3D / multi-plane imaging** | Reduces the orphan-nucleus problem caused by sectioning 3D cells into a 2D plane |
| **Post-hoc segmentation QC** | Check for biologically impossible co-expression (*CD3E* with *EPCAM*, *PTPRC* with *KRT*), inspect the fraction of unassigned transcripts, compare counts-per-cell against cell area, and **re-run with a second segmentation method** to see which conclusions survive. If a "novel cell state" disappears under a different segmentation, it was an artifact |
| **Segmentation-free analysis** | For some questions (spatial domains, gene co-localization, tissue niches) the cell is not required at all — analysing the transcript point cloud or binned data directly avoids the error entirely |

> **Practical rule:** never accept the vendor's default segmentation without looking at it overlaid on the image, and treat any surprising "hybrid" cell type as a segmentation artifact until proven otherwise.

### 7.4 Cell type annotation

| Aspect | scRNA-seq | Spatial |
|---|---|---|
| **Starting point** | A real single cell with whole-transcriptome coverage. Marker-based or reference-based annotation applies directly | Depends on family. Spot-based: the unit is a mixture, so the question shifts from *"which type is this?"* to *"which types compose this, in what proportion?"*. Imaging-based: the unit is a reconstructed cell, and annotation is only as trustworthy as its boundary |
| **Method** | Marker scoring, or label transfer from a reference (SingleR, Azimuth, scANVI, CellTypist) | Spot-based: **deconvolution** against a matched scRNA-seq reference (RCTD, Cell2location, SPOTlight). Imaging-based: same label-transfer logic as scRNA-seq, restricted to the panel genes |
| **Reference dependency** | Helpful, not mandatory | For spot-based, effectively **mandatory** — and the reference must match the tissue, condition and species. A mismatched reference will confidently assign cell types that are not there |
| **Gene coverage** | ~20,000 genes; any marker is available | Imaging platforms use a **closed panel** of a few hundred to a few thousand genes. If the markers distinguishing two subtypes were not designed into the panel, those subtypes are **unresolvable, permanently**. Panel design is therefore an annotation decision made before the experiment runs |
| **Achievable granularity** | Fine-grained states (e.g. Tem vs Trm vs exhausted CD8) | Spot-based: broad classes only; rare types get diluted below detection. Imaging-based: close to scRNA-seq granularity *if* the panel and segmentation cooperate — in practice, coarse labels are far more defensible than fine states |
| **Output** | One label per cell | Spot-based: a proportion vector per spot. Imaging-based: one label per segmented cell, plus a large fraction of low-count cells that resist confident labelling |
| **Validation** | Marker expression, known biology, reference agreement | All of the above **plus spatial coherence**, which is a genuinely independent check: annotated epithelium should line the lumen, germinal-centre B cells should form follicles, endothelium should trace vessels. Labels that are anatomically scattered are usually wrong |
| **Additional analyses unlocked** | None — position is destroyed at dissociation | Spatial domains, cellular niches and neighbourhoods, distance-to-structure analyses, and spatially constrained ligand-receptor inference |

> **Bottom line:** scRNA-seq annotation is essentially *"which reference, which markers"*. Spatial annotation carries a prior question that must be answered first — **"do I even trust my unit of measurement?"** For spot-based data that means asking how much cell-type mixing sits inside each spot; for imaging-based data it means asking how good the segmentation is. Skipping that question does not remove the error, it just makes it invisible.

---

## 8. Tools reference

- [Spatial_transcriptomics_tools](https://github.com/p-gueguen/Spatial_transcriptomics_tools) — curated list of tools for spatial transcriptomics analysis.


## Introduction to Single-cell Transcriptomics

# Learning objectives

By the end of this class, you should be able to:

- Recognize the main steps of single-cell and spatial transcriptomics analysis.
- Identify commonly used packages in R and Python.
- Understand which analytical problem is addressed at each step.
- Find reliable tutorials and best-practice recommendations for further study.

# One workflow, two ecosystems

R and Python provide complementary ecosystems for single-cell and spatial transcriptomics. Although package names and data structures differ, the underlying analytical workflow is largely the same.

| R ecosystem | Python ecosystem |
|---|---|
| Seurat and Bioconductor | Scanpy and scverse |
| `Seurat` or `SingleCellExperiment` objects | `AnnData` objects |
| Extensive statistical and visualization packages | Scalable workflows and machine-learning integration |

The choice of language should be guided by the biological question, the required methods, dataset size, available computing resources, and the experience of the research team.

# Single-cell RNA-seq analysis

A typical single-cell RNA-seq workflow transforms a gene-by-cell count matrix into biologically interpretable cell populations:

**Data import → Quality control → Normalization → Highly variable gene selection → Dimensionality reduction → Clustering → Cell-type annotation**

## 1. Data import

Single-cell analysis usually starts from a gene-by-cell matrix containing raw or filtered counts. Common inputs include Cell Ranger output directories, Matrix Market files (`.mtx`), HDF5 files (`.h5`), and processed `Seurat`, `SingleCellExperiment`, or `AnnData` objects.

| Task | R | Python |
|---|---|---|
| Read a 10x directory | `Seurat::Read10X()` | `scanpy.read_10x_mtx()` |
| Read a 10x HDF5 file | `Seurat::Read10X_h5()` | `scanpy.read_10x_h5()` |
| Main data structure | `Seurat` / `SingleCellExperiment` | `AnnData` |
| Common exchange formats | `.h5Seurat`, `.h5ad` | `.h5ad`, `.zarr` |

### R example

```{r}

library(Seurat)

counts <- Read10X("filtered_feature_bc_matrix/")
seu <- CreateSeuratObject(counts = counts)
```

### Python example

```{python}

import scanpy as sc

adata = sc.read_10x_mtx(
    "filtered_feature_bc_matrix/",
    var_names="gene_symbols"
)
```

## 2. Quality control

Quality control aims to remove empty droplets, damaged or dying cells, multiplets, and other technical artifacts before downstream analysis.

Frequently inspected cell-level metrics include:

- Total counts per cell.
- Number of detected genes per cell.
- Percentage of mitochondrial reads.
- Additional tissue-specific metrics, such as ribosomal or hemoglobin gene expression.

Thresholds should be adapted to the tissue, protocol, sequencing depth, and expected cell populations. Fixed universal cutoffs may remove biologically relevant cells or retain technical artifacts.

| Analytical task | R | Python |
|---|---|---|
| Calculate and visualize QC metrics | `Seurat`, `scater` | `Scanpy` |
| Identify empty droplets | `DropletUtils::emptyDrops()` | `CellBender` |
| Estimate ambient RNA contamination | `SoupX`, `DecontX` | `CellBender` |
| Detect doublets | `DoubletFinder`, `scDblFinder` | `Scrublet`, `SOLO` |

::: {.callout-important}
A high mitochondrial percentage or an unusually large number of detected genes is a warning signal, not an automatic reason to discard a cell. QC distributions should be evaluated for each dataset—and often for each sample.
:::

## 3. Normalization

Cells differ in sequencing depth, library size, and RNA capture efficiency. Normalization reduces these technical differences while preserving meaningful biological variation.

| Strategy | R | Python |
|---|---|---|
| Library-size normalization and log transformation | `Seurat::NormalizeData()` | `scanpy.pp.normalize_total()` followed by `scanpy.pp.log1p()` |
| Variance-stabilizing model | `Seurat::SCTransform()` | Pearson-residual-based methods |
| Deep generative model | — | `scvi-tools` |

### R example

```{r}

seu <- NormalizeData(seu)
```

### Python example

```{python}

sc.pp.normalize_total(adata, target_sum=1e4)
sc.pp.log1p(adata)
```

The appropriate strategy depends on the downstream analysis. Raw counts should be preserved because some statistical methods require them.

## 4. Highly variable gene selection

Not every detected gene is informative for distinguishing cellular states. Highly variable genes show greater cell-to-cell variation than expected from their average expression and are commonly used for dimensionality reduction and clustering.

| R | Python |
|---|---|
| `Seurat::FindVariableFeatures()` | `scanpy.pp.highly_variable_genes()` |
| `scran::modelGeneVar()` | Scanpy implementations of commonly used selection methods |

### R example

```{r}

seu <- FindVariableFeatures(
    seu,
    selection.method = "vst",
    nfeatures = 2000
)
```

### Python example

```{python}

sc.pp.highly_variable_genes(
    adata,
    n_top_genes=2000
)
```

Highly expressed genes are not necessarily highly variable genes. Feature selection focuses the analysis on genes that best describe differences among cells.

## 5. Dimensionality reduction

Single-cell datasets contain thousands of gene-expression dimensions. Dimensionality-reduction methods summarize this information using a smaller number of components.

| Method | Main purpose | R | Python |
|---|---|---|---|
| PCA | Summarize major sources of expression variation | `Seurat::RunPCA()` | `scanpy.tl.pca()` |
| UMAP | Visualize local and global cell relationships | `Seurat::RunUMAP()` | `scanpy.tl.umap()` |
| t-SNE | Visualize local cell relationships | `Seurat::RunTSNE()` | `scanpy.tl.tsne()` |

PCA is commonly used to construct the neighborhood graph and perform clustering. UMAP and t-SNE are primarily visualization methods.

::: {.callout-warning}
UMAP coordinates are visual representations, not direct biological measurements. Cluster size, shape, separation, and distance should not be overinterpreted without additional evidence.
:::

## 6. Graph construction and clustering

Cells are connected to transcriptionally similar neighbors to construct a graph. Community-detection algorithms then divide this graph into clusters.

| Task | R | Python |
|---|---|---|
| Construct the neighbor graph | `Seurat::FindNeighbors()` | `scanpy.pp.neighbors()` |
| Leiden clustering | `Seurat::FindClusters()` | `scanpy.tl.leiden()` |
| Identify cluster markers | `Seurat::FindAllMarkers()` | `scanpy.tl.rank_genes_groups()` |

The number and composition of clusters depend on choices such as the number of principal components, neighborhood size, integration strategy, and clustering resolution.

::: {.callout-important}
A cluster is a computational result—not automatically a cell type. Clusters must be interpreted and biologically validated.
:::

## 7. Cell-type annotation

Cell-type annotation assigns biological identities to computationally defined cells or clusters. Strong annotations combine multiple sources of evidence.

### Marker-based annotation

- Examine canonical and differentially expressed markers.
- Combine positive and negative marker evidence.
- Inspect marker co-expression rather than relying on a single gene.
- Consider tissue, disease, developmental stage, and experimental context.

### Reference-based annotation

| R | Python |
|---|---|
| `SingleR` | `CellTypist` |
| `Azimuth` | `scArches` |
| Seurat label transfer | `scvi-tools` |

Automated annotations depend on the quality and biological relevance of the reference. Predicted labels should therefore be treated as hypotheses and validated using marker expression and domain knowledge.

## Summary of the single-cell workflow

| Step | Main question |
|---|---|
| Data import | What data and metadata are available? |
| Quality control | Which droplets represent reliable cells? |
| Normalization | How can technical differences among cells be reduced? |
| Variable gene selection | Which genes capture informative variation? |
| Dimensionality reduction | What are the major expression patterns? |
| Clustering | Which cells have similar transcriptional profiles? |
| Annotation | What biological identities are supported by the evidence? |

# Spatial transcriptomics analysis

<!-- Spatial transcriptomics section to be developed by Dri. -->

# Recommended resources

- [Single-cell best practices](https://www.sc-best-practices.org/)
- [Orchestrating Single-Cell Analysis with Bioconductor](https://bioconductor.org/books/release/OSCA/)
- [Orchestrating Spatial Transcriptomics Analysis with Bioconductor](https://bioconductor.org/books/release/OSTA/)
- [Seurat guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
- [Scanpy tutorials](https://scanpy.readthedocs.io/en/stable/tutorials/)
- [scverse tutorials](https://scverse.org/learn/)

# Take-home message

R and Python provide different but complementary implementations of the same analytical workflow. A robust analysis depends more on informed methodological choices, transparent reporting, and biological validation than on the programming language itself.
